---
title: "Untitled"
author: "Yuchen Ma, Jianxiang Gao"
date: "2023-11-25"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Step 1: Data Preparation
## Loading and Inspecting Data

```{r}
# Load necessary libraries
library(readr)
library(dplyr)
library(tidyr)
library(readxl)

# Read the data from the CSV file
path1 <- "C:/Users/mayc0/OneDrive/桌面/Cancer_Rates.xlsx"
path2 <- "C:/Users/mayc0/OneDrive/桌面/Cancer_Rates.xlsx"
path3 <- "C:/Users/mayc0/OneDrive/桌面/Facility_Site_Points.xlsx"
path4 <- "C:/Users/mayc0/OneDrive/桌面/Obesity_Percentages.xlsx"
Cancer_Rates <- read_excel(path2)
facilities <- read_excel(path3)
obesity <- read_excel(path4)








# Inspect the structure of the data
str(Cancer_Rates)
glimpse(Cancer_Rates)
head(Cancer_Rates)

```

path1 <- "C:/Users/mayc0/OneDrive/桌面/Cancer_Rates.xlsx"
path2 <- "C:/Users/17404/Desktop/STAT 184/Cancer_Rates.csv"
path3 <- "C:/Users/17404/Desktop/STAT 184/Facility_Site_Points.csv"
path4 <- "C:/Users/17404/Desktop/STAT 184/Obesity_Percentages.csv"
Cancer_Rates <- read_csv(path2)
facilities <- read_csv(path3)
obesity <- read_csv(path4)


#Step 2: Data Wrangling
##Focus on selecting and cleaning the data relevant to the cancer types of interest.
```{r}
library(dplyr)
cancer_data_selected_raw <- select(Cancer_Rates, Colorectal, Lung_Bronc, Breast_Can, Prostate_C, Urinary_Sy)

# Round all numeric values to two decimal places
cancer_data_selected <- round(cancer_data_selected_raw, 2)

```
##Check for Missing Values:
```{r}
sum(is.na(cancer_data_selected))

```

#Step 3: Statistical Analysis
##Perform an ANOVA test to determine if there are differences among the cancer types.
```{r}
library(tidyr)
library(dplyr)

# Assuming cancer_data_selected is your dataframe with the relevant columns
cancer_data_long <- pivot_longer(
  cancer_data_selected,
  cols = c(Colorectal, Lung_Bronc, Breast_Can, Prostate_C, Urinary_Sy),
  names_to = "Cancer_Type",
  values_to = "Incidence_Rate"
)

# Now perform ANOVA with the correct formula
res <- aov(Incidence_Rate ~ Cancer_Type, data = cancer_data_long)
summary(res)

```


#Step 4: Data Visualization
##Create visualizations to support your statistical analysis.
```{r}
library(ggplot2)

ggplot(cancer_data_long, aes(x = Cancer_Type, y = Incidence_Rate)) +
    geom_boxplot() +
    labs(title = "Incidence Rates of Different Cancers in Lake County, Illinois", x = "Cancer Type", y = "Incidence Rate") +
    theme_classic() +
    theme(panel.grid.major = element_line(color = "gray", linetype = "dashed")) + 
  theme(plot.title = element_text(hjust = 0.5))


#Based on the ANOVA results quoted from your document, we have a highly significant F-value of 14.66 with a p-value of approximately 0. This indicates that there are statistically significant differences in the incidence rates of colorectal, lung, breast, prostate, and urinary system cancers in Lake County, Illinois. The extremely low p-value suggests that the likelihood that these differences could have occurred by chance is very small.

#To sum up, the data supports the existence of significant variability in the incidence rates of these cancers. This finding could serve as a basis for further investigation into the factors contributing to these differences, which could include genetic, environmental, lifestyle, or healthcare-related factors. The conclusion should also discuss the implications of these findings for public health policy, cancer research, and targeted interventions in Lake County, Illinois.






```

The follow codes aim to analyze the difference of incidence rate based on different zip code.

The result obtained is in the form of percentage, yet the % sign is omitted. i.e, 20 means 20%, -34 means -34%, or 34% lower than the median

Based on the result, clearly some of the regions has overall higher or lower incidence rate. Our next target is to try to analyze what caused the difference 
```{r}
library(kableExtra)
library(dplyr)
library(gt)


cancers_with_zip <- select(Cancer_Rates,ZIP, Colorectal, Lung_Bronc, Breast_Can, Prostate_C, Urinary_Sy) #now include zip code

difference_by_percentage <- function(value,median){
  return((value - median)/median)
  } #use to calculate the percentage away from the median

#this table store the values of medians of each cancer type
all_cancers_medians <- cancer_data_long %>% 
  group_by(Cancer_Type) %>%
  summarize(Median = median(Incidence_Rate)) 


#retrieve the value of each median
colorectal_median <- all_cancers_medians$Median[all_cancers_medians$Cancer_Type == "Colorectal"]
lung_median <- all_cancers_medians$Median[all_cancers_medians$Cancer_Type == "Lung_Bronc"]
Breast_median <- all_cancers_medians$Median[all_cancers_medians$Cancer_Type == "Breast_Can"]
Prostate_median <- all_cancers_medians$Median[all_cancers_medians$Cancer_Type == "Prostate_C"]
Urinary_median <- all_cancers_medians$Median[all_cancers_medians$Cancer_Type == "Urinary_Sy"]

#for each element, compute the percentage away from the corresponding median
cancers_difference_by_zip <- cancers_with_zip %>%
  mutate(Colorectal = difference_by_percentage(Colorectal,colorectal_median)*100,
         Lung_Bronc = difference_by_percentage(Lung_Bronc,lung_median)*100,
         Breast_Can = difference_by_percentage(Breast_Can,Breast_median)*100,
         Prostate_C = difference_by_percentage(Prostate_C,Prostate_median)*100,
         Urinary_Sy = difference_by_percentage(Urinary_Sy,Urinary_median)*100) %>%
  round(2)




#cancers_difference_by_zip %>%
#  kable(
#    caption = "Difference of incidence rate by zip",  
#    booktabs = TRUE,                        
#    align = c("l", rep("c", 6))        
#  ) %>%
# kableExtra::kable_styling(
#    bootstrap_options = c("striped"),
#    font_size = 16
#  )

#difference_by_zip_output <- cancers_difference_by_zip %>%
#  gt() %>%
#  tab_options(table.width = "100%") %>%
#  cols_label(
#    ZIP = "ZIP",
#    Colorectal = "Colorectal",
#    Lung_Bronc = "Lung",
#    Breast_Can = "Breast",
#    Prostate_C = "Prostate",
#    Urinary_Sy = "Urinary"
#  )

#difference_by_zip_output

#View(cancers_difference_by_zip)



```


Now, the focus should be on a few cases that is extremely abnormal. In this code chunk,
the average of variance of five cancers types has been calculated for each row, and
those zip codes with average higher or lower than 30 have been retained(i.e, in average, 
each cancer types is 30% higher or lower than the corresponding median)
```{r}
# Load the dplyr package
library(dplyr)



cancers_difference_by_zip$Average <- rowMeans(cancers_difference_by_zip[, c("Colorectal", "Lung_Bronc", "Breast_Can", "Prostate_C", "Urinary_Sy")], na.rm = TRUE)

# Create a new data frame with ZIP and the calculated average
cancers_difference_average <- data.frame(ZIP = cancers_difference_by_zip$ZIP, Average = cancers_difference_by_zip$Average)

abnormal_zip <- cancers_difference_average %>%
  filter(abs(Average)>20)

difference_by_zip_output <- abnormal_zip %>%
  gt() %>%
  tab_options(table.width = "100%") %>%
  cols_label(
    ZIP = "ZIP",
    Average = "Average"
  )

difference_by_zip_output

```

Obviously, four of the zipcodes has remarkably high cancer incidence rate, while the other three zipcodes has low incidence rate. There could be infinitely many causes, but the focus here are two factors: obesity rate and number of health facility
```{r}
health_facilities <- facilities %>%
  select(ZIP,FEATURECODE)%>%#only retain relevant rows
  filter(grepl("^(Hospital|Health|Medical)", FEATURECODE, ignore.case = TRUE)
         & grepl("^(60035|60040|60045|60069|60073|60085|60041)", ZIP))%>%
  mutate(ZIP = substr(ZIP, 1, 5))#We only interested in the first five digit of zipcode

head(health_facilities,10)

```
Yet, the health-related facilities in this data set are not inclusive and many of the entities has no recorded zip code, so we have only ten facilities related to corresponding zipcodes, which is not enough to make any conclusion. However, as you may notice, zipcode 60085 has five recorded health facilities, which could account for a reason of low cancer incidence rate(in average, 39.966% lower than the corresponding median) of that area.



Now we start looking at the obesity data
```{r}
obesity <- read_excel(path4)
excluded_zips <- c(60010,60020,60035, 60045, 60069, 60041)
# Filter out the specified zip codes
filtered_data <- obesity[!obesity$ZIP %in% excluded_zips, ]

# Compute mean values of Pct_Obese and ObsOrOvrwt for excluded zip codes
mean_excluded <- obesity[obesity$ZIP %in% excluded_zips, ] %>%
  summarise(Mean_Pct_Obese = mean(Pct_Obese), Mean_ObsOrOvrwt = mean(ObsOrOvrwt))

# Compute mean values for the remaining zip codes
mean_remaining <- filtered_data %>%
  summarise(Mean_Pct_Obese = mean(Pct_Obese), Mean_ObsOrOvrwt = mean(ObsOrOvrwt))

# Combine the results for comparison
results <- rbind(mean_excluded, mean_remaining)
row.names(results) <- c('Excluded Zip Codes', 'Remaining Zip Codes')
results

```
For the excluded zip codes:

Mean Percentage of Obese (Pct_Obese): 0.1995714 (approximately 19.96%)
Mean Percentage of Overweight or Obese (ObsOrOvrwt): 0.5452857 (approximately 54.53%)
For the remaining zip codes:

Mean Percentage of Obese (Pct_Obese): 0.2279500 (approximately 22.80%)
Mean Percentage of Overweight or Obese (ObsOrOvrwt): 0.5718500 (approximately 57.19%)


```{r}
selected_zip_codes <- c(60010, 60020, 60035, 60045, 60069, 60041)

filtered_cancer <- Cancer_Rates[Cancer_Rates$ZIP %in% selected_zip_codes, ]

selected_data <- filtered_cancer$All_Cancer
```



```{r}
excluded_data <- c(60010, 60020, 60035, 60045, 60069, 60041)
filtered_obse <- obesity[obesity$ZIP %in% excluded_data, ]
```





```{r}
library(ggplot2)

ggplot(data = filtered_obse, aes(x = Pct_Obese, y = selected_data)) +
    geom_point() +
      geom_smooth(method = "lm", color = "yellow") +

    labs(x = "Percentage of Obese",
         y = "Incidence Rate",
         title = "Scatter plot of Percentage of Obese vs. Incidence Rate") +
    theme_minimal()

```


```{r}
ggplot(data = filtered_obse, aes(x = ObsOrOvrwt, y = selected_data)) +
    geom_point() +
        geom_smooth(method = "lm", color = "green") +

    labs(x = "Percentage of Overweight and Obese",
         y = "Incidence Rate",
         title = "Scatter plot of Percentage of Overweight and Obese vs. Incidence Rate") +
    theme_minimal()

```



















Then for rigorous research, we conduct t test.

```{r}
library(broom)

# Separate the dataset into two groups
excluded_data <- obesity %>% filter(ZIP %in% excluded_zips)
remaining_data <- obesity %>% filter(!ZIP %in% excluded_zips)

# Perform a t-test for Pct_Obese
t_test_obese <- t.test(excluded_data$Pct_Obese, remaining_data$Pct_Obese)

# Perform a t-test for ObsOrOvrwt
t_test_overweight <- t.test(excluded_data$ObsOrOvrwt, remaining_data$ObsOrOvrwt)

# Output the results of the t-tests
list(
  Pct_Obese = tidy(t_test_obese),
  ObsOrOvrwt = tidy(t_test_overweight)
)

```
Here's the summary of the results:
For the Pct_Obese variable:

Estimate of difference (mean in excluded zip codes - mean in remaining zip codes): -0.02837857
p-value: 0.4086131
For the ObsOrOvrwt variable:

Estimate of difference: -0.02656429
p-value: 0.476115
The p-values in both tests are greater than 0.05, indicating that there are no statistically significant differences in the mean percentages of obesity (Pct_Obese) and overweight or obese (ObsOrOvrwt) between the excluded zip codes and the remaining zip codes.




