---
title: "Cancers Incidence Rate Analysis on Lake County, Illinois"
author: "Yuchen Ma, Jianxiang Gao"
date: "2023-11-25"
output:
  html_document: default
  pdf_document:
    latex_engine: xelatex
  word_document: default
---

```{r setup, include=FALSE, message = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

In my consciousness, I believed the incidence rate of cancers should be uniform in a relatively large region. I never think about it in more depth, until I had a conversation with my friend living in Lake County, Illinois. He told me many people he know living in his community got cancers, and that drive me to think: is it possible that the cancer incidence rates actually vary in a small area scope, for example, zip codes?

This lead me to raise this research question:
Is there a statistically significant difference in the incidence rates of five cancer types among different areas(ZIP code) of Lake County, Illinois? If so, what cause it?


## Data Inspectation and Wrangling

We have analyzed three data sets to answer our research question, which are: 'cancer_rate', 'facilities', and 'obesity'. All data are published by government of Lake County, Illinois. All data are already in tidy-data form. The following columns is what we used in our analysis.

cancer_rate:
The data in columns 2-7 are average incidence rate per 100,000 people

1. ZIP
2. Colorectal Cancer
3. Lung Cancer
4. Breast Cancer
5. Prostate Cancer
6. Urinary System Cancer
7. All Cancer: Incidence rate of all reported cancers(included but not limited to above five cancer types ).

facilities:
1.ZIP
2.Feature Code: type of the facilities(ex: Law Enforcement, Hospital, College, etc.).

obesity:
1.ZIP
2.Overweight percentage(Including Obese)
3.Obese percentage

Data Wrangling process in this project only contain some selection and structure conversion as the original data are already in tidy form.


```{r message = FALSE}
# The programming style we used is Snake Case
library(readr)
library(dplyr)
library(tidyr)
library(readxl)

path1 <- "C:/Users/mayc0/OneDrive/桌面/Cancer_Rates.xlsx"
path2 <- "C:/Users/mayc0/OneDrive/桌面/Cancer_Rates.xlsx"
path3 <- "C:/Users/mayc0/OneDrive/桌面/Facility_Site_Points.xlsx"
path4 <- "C:/Users/mayc0/OneDrive/桌面/Obesity_Percentages.xlsx"



cancer_rate <- read_excel(path2)
facilities <- read_excel(path3)
obesity <- read_excel(path4)







```

path1 <- "C:/Users/mayc0/OneDrive/桌面/Cancer_Rates.xlsx"
path2 <- "C:/Users/17404/Desktop/STAT 184/Cancer_Rates.csv"
path3 <- "C:/Users/17404/Desktop/STAT 184/Facility_Site_Points.csv"
path4 <- "C:/Users/17404/Desktop/STAT 184/Obesity_Percentages.csv"
cancer_rate <- read_csv(path2)
facilities <- read_csv(path3)
obesity <- read_csv(path4)

```{r message = FALSE}
library(dplyr)
cancer_data_selected_raw <- select(cancer_rate, Colorectal, Lung_Bronc, Breast_Can, Prostate_C, Urinary_Sy)

# Round all numeric values to two decimal places
cancer_data_selected <- round(cancer_data_selected_raw, 2)

```


##Statistical Analysis of Variance Existence
Before analyzing difference in the incidence rates of five cancer types among different ZIP code, we have to answer one question. Is there a variance of incidence between five cancer types? To answer this, we performed ANOVA test. Our Null Hypothesis is: there is no difference between incidence rates of five cancer types

```{r message = FALSE}
library(tidyr)
library(dplyr)

# Assuming cancer_data_selected is your dataframe with the relevant columns
cancer_data_long <- pivot_longer(
  cancer_data_selected,
  cols = c(Colorectal, Lung_Bronc, Breast_Can, Prostate_C, Urinary_Sy),
  names_to = "Cancer_Type",
  values_to = "Incidence_Rate"
)

# Now perform ANOVA with the correct formula
res <- aov(Incidence_Rate ~ Cancer_Type, data = cancer_data_long)
summary(res)

```

Based on the ANOVA results, we have a highly significant F-value of 14.66 with a p-value of approximately 0. This indicates that there are statistically significant differences in the incidence rates of colorectal, lung, breast, prostate, and urinary system cancers in Lake County, Illinois. The extremely low p-value suggests that the likelihood that these differences could have occurred by chance is very small.

The figure below displays a box plot of incidence rate of five cancers type, each point represent the incidence rate of cancer type of a ZIP code
```{r message = FALSE}
library(ggplot2)

ggplot(cancer_data_long, aes(x = Cancer_Type, y = Incidence_Rate)) +
    geom_boxplot() +
    labs(title = "Incidence Rates of Different Cancers in Lake County, Illinois", x = "Cancer Type", y = "Incidence Rate") +
    theme_classic() +
    theme(panel.grid.major = element_line(color = "gray", linetype = "dashed")) + 
  theme(plot.title = element_text(hjust = 0.5))









```
Apparently, Breast Cancer has a remarkably high incidence rates, while the other four types have a similar incidence rates.


In order answer the research question, we need to calculate the variance of average incidence rate away from the median for each ZIP code. We do this by calculating the average of variance of five cancers types, and those zip codes with average variance higher or lower than 20 have been retained(i.e, in average, each cancer types is 30% higher or lower than the corresponding median)


Based on the bar graph below, clearly some of the regions has overall higher or lower incidence rate. Our next target is to try to analyze what caused the difference 
```{r message = FALSE}
library(dplyr)
library(gt)


cancers_with_zip <- select(cancer_rate,ZIP, Colorectal, Lung_Bronc, Breast_Can, Prostate_C, Urinary_Sy) #now include zip code

difference_by_percentage <- function(value,median){
  return((value - median)/median)
  } #use to calculate the percentage away from the median

#this table store the values of medians of each cancer type
all_cancers_medians <- cancer_data_long %>% 
  group_by(Cancer_Type) %>%
  summarize(Median = median(Incidence_Rate)) 


#retrieve the value of each median
colorectal_median <- all_cancers_medians$Median[all_cancers_medians$Cancer_Type == "Colorectal"]
lung_median <- all_cancers_medians$Median[all_cancers_medians$Cancer_Type == "Lung_Bronc"]
Breast_median <- all_cancers_medians$Median[all_cancers_medians$Cancer_Type == "Breast_Can"]
Prostate_median <- all_cancers_medians$Median[all_cancers_medians$Cancer_Type == "Prostate_C"]
Urinary_median <- all_cancers_medians$Median[all_cancers_medians$Cancer_Type == "Urinary_Sy"]

#for each element, compute the percentage away from the corresponding median
cancers_difference_by_zip <- cancers_with_zip %>%
  mutate(Colorectal = difference_by_percentage(Colorectal,colorectal_median)*100,
         Lung_Bronc = difference_by_percentage(Lung_Bronc,lung_median)*100,
         Breast_Can = difference_by_percentage(Breast_Can,Breast_median)*100,
         Prostate_C = difference_by_percentage(Prostate_C,Prostate_median)*100,
         Urinary_Sy = difference_by_percentage(Urinary_Sy,Urinary_median)*100) %>%
  round(2)




#cancers_difference_by_zip %>%
#  kable(
#    caption = "Difference of incidence rate by zip",  
#    booktabs = TRUE,                        
#    align = c("l", rep("c", 6))        
#  ) %>%
# kableExtra::kable_styling(
#    bootstrap_options = c("striped"),
#    font_size = 16
#  )

#difference_by_zip_output <- cancers_difference_by_zip %>%
#  gt() %>%
#  tab_options(table.width = "100%") %>%
#  cols_label(
#    ZIP = "ZIP",
#    Colorectal = "Colorectal",
#    Lung_Bronc = "Lung",
#    Breast_Can = "Breast",
#    Prostate_C = "Prostate",
#    Urinary_Sy = "Urinary"
#  )

#difference_by_zip_output

#View(cancers_difference_by_zip)



```


```{r message = FALSE}
# Load the dplyr package
library(dplyr)
library(ggplot2)



cancers_difference_by_zip$Average <- rowMeans(cancers_difference_by_zip[, c("Colorectal", "Lung_Bronc", "Breast_Can", "Prostate_C", "Urinary_Sy")], na.rm = TRUE)

# Create a new data frame with ZIP and the calculated average
cancers_difference_average <- data.frame(ZIP = cancers_difference_by_zip$ZIP, Average = cancers_difference_by_zip$Average)

abnormal_zip <- cancers_difference_average %>%
  filter(abs(Average)>20)

#difference_by_zip_output <- abnormal_zip %>%
#  gt() %>%
#  tab_options(table.width = "100%") %>%
#  cols_label(
#    ZIP = "ZIP",
#    Average = "Average"
#  )

abnormal_zip$ZIP <- factor(abnormal_zip$ZIP)

ggplot(
  data = abnormal_zip,
  aes(x = ZIP, y = Average)
) +
  geom_bar(stat = "identity", fill = "steelblue",width = 0.7) +  # Adjust fill color as needed
  theme_minimal() +  # You can change the theme as per your preference
  labs(title = "Average variance from the median", 
       x = "ZIP", y = "Average Variance") + theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5),    
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )

```

## Statistical Analysis of Cause of The Variance

There could be infinitely many causes, but the focus here are two factors: obesity rate and number of health facility

### Statistical Analysis of Health Facility

```{r message = FALSE}
health_facilities <- facilities %>%
  select(ZIP,FEATURECODE)%>%#only retain relevant rows
  filter(grepl("^(Hospital|Health|Medical)", FEATURECODE, ignore.case = TRUE)
         & grepl("^(60010|60020|60035|60045|60069|60041|60040|60064|60073|60085)", ZIP))%>%
  mutate(ZIP = substr(ZIP, 1, 5))#We only interested in the first five digit of zipcode

head(health_facilities,30)

```
Yet, the health-related facilities in this data set are not inclusive and many of the entities has no recorded zip code, so we have only ten facilities related to corresponding zipcodes, which is not enough to make any conclusion. However, as you may notice, zipcode 60085 has five recorded health facilitwies, which could account for a reason of low cancer incidence rate(in average, 39.966% lower than the corresponding median) of that area.


### Statistical Analysis of Obesity Rate



#### Intro to Obesity Data

Now we start looking at the obesity data
First let's take a look at obesity data. This data has two variables, they are:

Pct_Obese – The percent of people in the zip code who are considered obese, defined as having a BMI greater than or equal to 30.

ObsOrOvrwt –The percent of people in the zip code who are considered overweight (defined as having a BMI greater than or equal to 25 but less than 30) or obese (defined as having a BMI greater than or equal to 30).


Then we selected six of the above ten abnormal zipcodes with positive variances for testing convenience.




#### Mean

Next, we compare the average obesity data of these six abnormal zipcodes with the average obesity data of other zipcodes.
```{r message = FALSE}
obese_average <- mean(obesity$Pct_Obese)
overweight_average <- mean(obesity$ObsOrOvrwt)

excluded_zips <- c( 60010,60020,60035,60045, 60069,60041)
# Filter out the specified zip codes
filtered_data <- obesity[!obesity$ZIP %in% excluded_zips, ]

# Compute mean values of Pct_Obese and ObsOrOvrwt for excluded zip codes
mean_excluded <- obesity[obesity$ZIP %in% excluded_zips, ] %>%
  summarise(Mean_Pct_Obese = mean(Pct_Obese), Mean_ObsOrOvrwt = mean(ObsOrOvrwt))

# Compute mean values for the remaining zip codes
mean_remaining <- filtered_data %>%
  summarise(Mean_Pct_Obese = mean(Pct_Obese), Mean_ObsOrOvrwt = mean(ObsOrOvrwt))

# Combine the results for comparison
results <- rbind(mean_excluded, mean_remaining)
row.names(results) <- c('Excluded Zip Codes', 'Remaining Zip Codes')
results

```
For the excluded zip codes:

Mean Percentage of Obese (Pct_Obese): 0.17 (approximately 17.00%)
Mean Percentage of Overweight or Obese (ObsOrOvrwt): 0.5111667 (approximately 51.12%)
For the remaining zip codes:

Mean Percentage of Obese (Pct_Obese): 0.2350476 (approximately 23.50%)
Mean Percentage of Overweight or Obese (ObsOrOvrwt): 0.5803333 (approximately 58.03%)

We found that there was a difference but not a huge one.




#### t-test


Then for rigorous research, we conduct t-test.
Hypothesis test:
Is there a difference between obesity rate of the six abnormal zipcodes and obesity rate of other zipcodes?"

the hypotheses might be formally stated as:
Null hypothesis: There is no difference in the obesity rate between the six abnormal ZIP codes and the other ZIP codes.
Alternative hypothesis: There is a difference in the obesity rate between the six abnormal ZIP codes and the other ZIP codes.
We assume that the significance level is 5%.

```{r message = FALSE}
library(broom)

# Separate the dataset into two groups
excluded_data <- obesity %>% filter(ZIP %in% excluded_zips)
remaining_data <- obesity %>% filter(!ZIP %in% excluded_zips)

# Perform a t-test for Pct_Obese
t_test_obese <- t.test(excluded_data$Pct_Obese, remaining_data$Pct_Obese)

# Perform a t-test for ObsOrOvrwt
t_test_overweight <- t.test(excluded_data$ObsOrOvrwt, remaining_data$ObsOrOvrwt)

# Output the results of the t-tests
list(
  Pct_Obese = tidy(t_test_obese),
  ObsOrOvrwt = tidy(t_test_overweight)
)

```



#### P-value
We mainly care about the p-value, so we specifically extracted the p-values of the two sets of t-tests and made them into tables.

```{r}
p_values <- data.frame(
  Test = c("Pct_Obese", "ObsOrOvrwt"),
  P_Value = c(t_test_obese$p.value, t_test_overweight$p.value)
)

ggplot(p_values, aes(x = Test, y = P_Value)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red") +
  annotate("text", x = 1.5, y = 0.055, label = "Significance threshold (0.05)", color = "red", size = 3.5) +
  geom_text(aes(label = round(P_Value, 3)), vjust = -0.5) +
  ylim(0, 0.075) +
  labs(title = "P-Values of T-Tests", x = "Test", y = "P-Value") +
  theme_minimal()

```



#### scatter plots

In order to show the relationship between obesity and cancer incidence rate more intuitively, we related the cancer incidence rate of the six zipcodes to the Percentage of Overweight or Obese and the Percentage of Obese respectively, and drew two scatter plots.

```{r message = FALSE}
excluded_zips <- c(60010, 60020, 60035, 60045, 60069, 60041)
filtered_cancer <- cancer_rate[cancer_rate$ZIP %in% excluded_zips, ]
selected_data <- filtered_cancer$All_Cancer
filtered_obse <- obesity[obesity$ZIP %in% excluded_zips, ]

#weight_vs_incidence <- merge(filtered_obse, cancer_rate, by = "ZIP", all.x = TRUE, all.y = TRUE) %>%
#  select(ZIP, Pct_Obese, ObsOrOvrwt, All_Cancer)

ggplot(data = filtered_obse, aes(x = ObsOrOvrwt, y = selected_data)) +
    geom_point() +
        geom_smooth(method = "lm", color = "green") +

    labs(x = "Percentage of Overweight or Obese",
         y = "Incidence Rate",
         title = "Scatter plot of Percentage of Overweightor or Obese vs. Incidence Rate") +
    theme_minimal()+
    theme(plot.title = element_text(hjust = 0.5))

```


```{r message = FALSE}
ggplot(data = filtered_obse, aes(x = Pct_Obese, y = selected_data)) +
    geom_point() +
      geom_smooth(method = "lm", color = "yellow") +

    labs(x = "Percentage of Obese",
         y = "Incidence Rate",
         title = "Scatter plot of Percentage of Obese vs. Incidence Rate") +
    theme_minimal()+
    theme(plot.title = element_text(hjust = 0.5))
```




#### Test Summary

Here's the summary of the results:
For the ObsOrOvrwt variable:
p-value: 0.043

For the Pct_Obese variable:
p-value: 0.042

The p-values in both tests are greater than 0.05 and we reject the null hypothesis, indicating that there are statistically significant differences in the mean percentages of obesity (Pct_Obese) and overweight or obese (ObsOrOvrwt) between the excluded zip codes and the remaining zip codes.

This shows that in areas with abnormal cancer incidence rates such as 60010, 60020, 60035, 60045, 60069, 60041, the probability of obesity and overweight or obese is significantly different from areas other than areas with abnormal cancer incidence rates.

Therefore, based on the conclusion of t-test, we boldly speculate that the statistically significant difference in the incidence rates of five cancer types among different areas (ZIP code) of Lake County, Illinois is caused by obesity.

But in fact we can only guess that the obesity rate may affect the cancer incidence rate, because our t-test can only show that the obesity rate is also abnormal in other areas in places where cancer is abnormal, but we have not yet determined the reasons for the difference.






# Conclusion 

First, we found and proved through ANOVA test and box plot that there is a statistically significant difference in the incidence rates of five cancer types among different areas (ZIP code) of Lake County, Illinois. 
Then, we calculate the variance of average incidence rate away from the median for each ZIP code and drew a bar graph, and found 10 zipcodes with abnormal cancer incidence rates.

After that, we started the second phase of the project, finding the reasons for the abnormal cancer rates. Due to endless factors, we chose to study the two most important factors: obesity rate and number of health facilities.
In the study on number of health facilities, we were unable to draw conclusions because there were too few data. But we found that zipcode 60085 has five recorded health facilities, which could account for a reason of low cancer incidence rate.
Finally, in our study of obesity rates, we used t-tests and graphs to find the obesity rate is also abnormal in other areas in places where cancer incidence rate is abnormal, so we can only guess that the obesity rate may affect the cancer incidence rate, but we have not yet determined the reasons for the difference.


# Appendix: All code for this report

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```


